---
title: "Comprehensive Analysis of GC-IMS Features for Cohort Classification P1.2"
author: "Deevanshi Walia"
date: "`r Sys.Date()`"
output:
  word_document:
    toc: yes
    toc_depth: 3
---

```{r setup, include=FALSE}
# This chunk sets up the global options for the document.
# include=FALSE means this chunk won't appear in the final report.
# echo=FALSE means the code itself won't be shown, just the output.
# message=FALSE and warning=FALSE prevent messages and warnings from appearing.
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)


library(tools)
library(dbscan)
library(ggplot2)
library(reshape2)
library(dplyr); library(readr); library(caret); library(tidyr); library(MASS)
library(pROC); library(ggplot2); library(randomForest); library(e1071)
library(xgboost); library(GGally); library(tibble)
```
# 1. Data Loading and Preparation

This section details the loading and initial preparation of the feature matrix that serves as the input for all subsequent analyses.

```{r load_data}
# Load libraries for this chunk
library(readr)
library(dplyr)
library(tibble)

feature_matrix_path <-"C:/Users/deevanshi.walia/Desktop/Try 2.0/CT-Value_25CT/Final_Analysis_Output_Peaks_Clustering_9_0.08/Feature_Matrices/feature_matrix_MEDIAN_INTENSITY.csv"
data_full<- read_csv(feature_matrix_path, show_col_types = FALSE)
data_full$GroupName <- as.factor(data_full$GroupName)
if (any(is.na(data_full))) {
  warning("NA values detected. Removing rows with NAs.")
  data_full <- na.omit(data_full)
}
metadata <- data_full %>% dplyr:: select(SampleName, GroupName)
feature_data <- data_full %>% dplyr::select(starts_with("Cluster_"))
```

The dataset was successfully loaded, containing `r nrow(feature_data)` samples and `r ncol(feature_data)` features.

# 2. Robust Feature Selection with Boruta

To identify a minimal, non-redundant, and predictively powerful subset of features, the Boruta algorithm was employed.

```{r run_boruta}
# Load libraries for this chunk
library(Boruta)
library(dplyr)
library(vegan)
library(MASS)
library(randomForest) 
library(e1071)
library(xgboost)
library(pROC)
library(caret)
# --- STAGE 1: Broad Screen with Boruta ---
set.seed(456)
boruta_result <- Boruta(x = feature_data, y = metadata$GroupName, doTrace = 2, maxRuns = 700)
boruta_selected_features <- getSelectedAttributes(boruta_result, withTentative = FALSE)
```
The Boruta algorithm confirmed **`r length(boruta_selected_features)`** features as statistically important. The selected features are: `r paste(boruta_selected_features, collapse=", ")`.

# 3. Statistical Confirmation with Wilcoxon Test

```{r Wilcoxon Test on the selected features}
if (length(boruta_selected_features) > 0) {
  univariate_results_df <- lapply(boruta_selected_features, function(feature_name) {
    test_result <- wilcox.test(feature_data[[feature_name]] ~ metadata$GroupName)
    data.frame(Feature = feature_name, p.value = test_result$p.value)
  }) %>%
    bind_rows()
  
  SIGNIFICANCE_THRESHOLD <- 0.05
  selected_features <- univariate_results_df %>%
    filter(p.value < SIGNIFICANCE_THRESHOLD) %>%
    pull(Feature)
  
} 

if (length(selected_features) == 0) {
  stop("The two-stage selection process resulted in zero features.")
}

feature_data_best <- data_full %>% dplyr::select(all_of(selected_features))
```
The algorithm selected the following features `r boruta_selected_features`.
The final, high-confidence signature (Boruta-confirmed AND p < 0.05) is:
`r print(selected_features)`

```{r plot_boruta_ggplot, fig.width=12, fig.height=7}
library(ggplot2)
library(dplyr)
library(tidyr) 
library(forcats) 

boruta_plot_data <- as.data.frame(boruta_result$ImpHistory) %>%
  # Remove the shadow feature columns from the main data for clarity
  dplyr::select(-matches("shadow")) %>%
  pivot_longer(
    cols = everything(),
    names_to = "Feature",
    values_to = "Importance"
  )

# --- Step 2: Get the Final Decision for Each Feature for Coloring ---
feature_decisions <- boruta_result$finalDecision %>%
  tibble::enframe(name = "Feature", value = "Decision")

# Join the decisions back to the plot data
boruta_plot_data <- boruta_plot_data %>%
  left_join(feature_decisions, by = "Feature") %>%
  # Filter out any rejected features to keep the plot clean
  filter(Decision != "Rejected") %>%
  # Reorder the features on the x-axis by their median importance
  mutate(Feature = fct_reorder(Feature, Importance, .fun = median))

# --- Step 3: Create the ggplot ---
boruta_ggplot <- ggplot(
  data = boruta_plot_data,
  aes(x = Feature, y = Importance, fill = Decision)
) +
  # Add the boxplots
  geom_boxplot(alpha = 0.8, outlier.shape = 21, outlier.size = 2) +
  
  # Define the colors for the decisions
  scale_fill_manual(
    name = "Boruta Decision",
    values = c("Confirmed" = "darkgreen", "Tentative" = "darkorange")
  ) +
  
  # Add a horizontal line at y=0 for reference
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  
  labs(
    title = "Boruta Feature Importance",
    subtitle = "Feature importance across all iterations",
    x = "Features",
    y = "Importance"
  ) +
  
  coord_flip() +
  
  theme_bw(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5),
    legend.position = "bottom",
    axis.text.y = element_text(size = 10)
  )
boruta_ggplot
```


# 4. Model Gauntlet
```{r model_gauntlet, echo=FALSE, message=FALSE, warning=FALSE}
library(vegan); library(MASS); library(randomForest); library(e1071);
library(xgboost); library(pROC); library(caret); library(tibble)

num_samples <- nrow(feature_data_best)
y_response <- as.factor(metadata$GroupName)
y_response_numeric <- as.numeric(y_response) - 1
current_group_levels <- levels(y_response)
positive_class_name <- current_group_levels[1]
loocv_results <- data.frame(SampleName = metadata$SampleName, ActualGroup = y_response,
                            lda_scores=numeric(num_samples), rf_scores=numeric(num_samples),
                            svm_scores=numeric(num_samples), xgb_scores=numeric(num_samples))


for (i in 1:num_samples) {
  x_train <- as.matrix(feature_data_best[-i, , drop=FALSE]); y_train <- y_response[-i]
  y_train_numeric <- y_response_numeric[-i]; x_test <- as.matrix(feature_data_best[i, , drop = FALSE])
  
  lda_model <- lda(x = x_train, grouping = y_train)
  loocv_results$lda_scores[i] <- predict(lda_model, newdata = data.frame(x_test))$posterior[, positive_class_name]
  
  min_class_size <- min(table(y_train))
  rf_model <- randomForest(x = x_train, y = y_train, ntree = 500, sampsize = c(min_class_size, min_class_size))
  loocv_results$rf_scores[i] <- predict(rf_model, newdata = x_test, type = "prob")[, positive_class_name]
  
  svm_model <- svm(x = x_train, y = y_train, probability = TRUE, kernel="radial")
  loocv_results$svm_scores[i] <- attr(predict(svm_model, newdata = data.frame(x_test), probability = TRUE), "probabilities")[, positive_class_name]
  
  dtrain <- xgb.DMatrix(data = x_train, label = y_train_numeric)
  dtest <- xgb.DMatrix(data = x_test)
  scale_pos_weight <- sum(y_train_numeric == 0) / sum(y_train_numeric == 1)
  params <- list(objective="binary:logistic", eval_metric="logloss", scale_pos_weight=scale_pos_weight)
  xgb_model <- xgboost(params = params, data = dtrain, nrounds = 50, verbose = 0)
  loocv_results$xgb_scores[i] <- predict(xgb_model, dtest)
}

roc_lda <- roc(response = loocv_results$ActualGroup, predictor = loocv_results$lda_scores, levels=current_group_levels)
roc_rf  <- roc(response = loocv_results$ActualGroup, predictor = loocv_results$rf_scores, levels=current_group_levels)
roc_svm <- roc(response = loocv_results$ActualGroup, predictor = loocv_results$svm_scores, levels=current_group_levels)
roc_xgb <- roc(response = loocv_results$ActualGroup, predictor = loocv_results$xgb_scores, levels=current_group_levels)

model_performance_list <- list()
models_to_test <- list(LDA = roc_lda, `Random Forest` = roc_rf, SVM = roc_svm, XGBoost = roc_xgb)
cm_list <- list() 
for (model_name in names(models_to_test)) {
  current_roc <- models_to_test[[model_name]]
  best_coords <- coords(current_roc, "best", ret = c("threshold", "youden"), best.method="youden")
  best_thresh <- best_coords$threshold; youdens_j <- best_coords$youden
  predicted_classes <- ifelse(current_roc$predictor >= best_thresh, current_group_levels[2], current_group_levels[1])
  cm_stats <- confusionMatrix(data = factor(predicted_classes, levels = current_group_levels), reference = current_roc$response, positive = positive_class_name)
  cm_list[[model_name]] <- cm_stats
  model_performance_list[[model_name]] <- c(AUC=auc(current_roc), cm_stats$overall["Accuracy"], 
cm_stats$byClass[c("Sensitivity", "Specificity", "Balanced Accuracy")], "Youden_J"=youdens_j,"Pos Pred Value","Neg Pred Value")
}
performance_summary <- as.data.frame(do.call(rbind, model_performance_list)) %>%
  tibble::rownames_to_column("Model") %>%
  mutate(across(where(is.numeric), ~ round(., 4))) %>%
  dplyr::select(Model, AUC, Accuracy, `Balanced Accuracy`, Youden_J, Sensitivity, Specificity)
```
### 4.1. Comparative Model Performance

The table below summarizes the performance of four different classification models trained on the selected features and validated with Leave-One-Out Cross-Validation.


```{r print_performance_summary, echo=FALSE, message=FALSE, warning=FALSE}
# Using kable() on the performance summary table
knitr::kable(performance_summary, caption = "Comparative performance metrics for all tested classifiers.")
```

### 4.2. ROC Curve and Best Model Analysis

```{r plot_roc_and_cm}

roc_list_for_plot <- list(LDA = roc_lda, RF = roc_rf, SVM = roc_svm, XGBoost = roc_xgb)
roc_plot_combined <- ggroc(roc_list_for_plot, size = 1) +
  annotate("segment", x = 1, xend = 0, y = 0, yend = 1, color="grey", linetype="dashed") +
  labs(title = "Comparative ROC Curves for All Models", color = "Model") +
  theme_minimal(base_size = 14) + theme(plot.title = element_text(face = "bold", hjust = 0.5))
roc_plot_combined

best_model_name <- performance_summary$Model[which.max(performance_summary$`Balanced Accuracy`)]
cat("\n--- Detailed Confusion Matrix for the Best Performing Model:", best_model_name, "---\n")

# Get the confusion matrix object for the best model
best_cm_object <- cm_list[[best_model_name]]

# --- 1. Create the main confusion matrix table ---
cm_table <- as.data.frame(best_cm_object$table)
colnames(cm_table) <- c("Prediction", "Reference", "Frequency")
cm_table_wide <- tidyr::pivot_wider(cm_table, names_from = "Reference", values_from = "Frequency")

# Use kable() to format it nicely
knitr::kable(
  cm_table_wide,
  caption = paste("Confusion Matrix for the", best_model_name, "Model")
)
# --- 2. Create a separate table for the key statistics ---
# Extract the key metrics we want to show
key_stats <- c(
  "Accuracy",
  "Kappa",
  "Balanced Accuracy",
  "Sensitivity",
  "Specificity",
  "Pos Pred Value",
  "Neg Pred Value"
)

# Combine overall and by-class stats
all_stats <- c(best_cm_object$overall, best_cm_object$byClass)

# Create a clean data frame of the stats
stats_to_display <- data.frame(
  Metric = key_stats,
  Value = round(all_stats[key_stats], 4) # Round to 4 decimal places
)

# Use kable() to format this table as well
knitr::kable(
  stats_to_display,
  caption = "Key Performance Statistics for the Best Model"
)
```

# 5. Intensity Distributions
```{r feature interactions, fig.width=12, fig.height=12}
library(ggpubr)
library(GGally)
library(ggplot2)
# A pairs plot is most useful if you have between 2 and ~6 features.
if (length(selected_features) >= 2) {
  
  # 1. Select the data for the plot
  # This includes the GroupName and all the selected cluster intensities
  data_for_pairs_plot <- cbind(
    GroupName = metadata$GroupName,
    feature_data[, selected_features]
  )

# --- Create the pairs plot with improved theme settings ---
pairs_plot <- ggpairs(
  data_for_pairs_plot,
  mapping = aes(color = GroupName, alpha = 0.7),
  columns = 2:ncol(data_for_pairs_plot),
  lower = list(continuous = wrap("points", size = 2)),
  diag = list(continuous = wrap("densityDiag", alpha = 0.5)),
  upper = list(continuous = "blank")
) +
  labs(title = paste("Pairwise Interactions of", length(selected_features), "Selected Features")) +
  
  theme_bw(base_size = 10) + 
  theme(
    # Center the main plot title
    plot.title = element_text(face="bold", size=16, hjust=0.5),
    
    # Adjust the panel strip text (the "Cluster_X" labels at the top and right)
    strip.text.x = element_text(
      size = 10, # Make the font smaller
      angle = 0, # Angle the text to prevent overlap
      hjust = 0.5,# Adjust horizontal alignment
      face= "bold" 
    ),
    strip.text.y = element_text(
      size = 10,
      angle = 90, # Keep y-axis labels horizontal
      hjust = 0.5, # Adjust horizontal alignment
      face="bold"   
    ),
    
    # Give the axis labels a bit more room
    axis.text.x = element_text(size=7, angle=45, vjust=0.5),
    axis.text.y = element_text(size=7),
    
    # Move the legend to the bottom to free up space on the right
    legend.position = "bottom"
  )

# This will now embed the cleaner, more readable plot
pairs_plot
 
} else{
  print("   - Skipping pairs plot: Fewer than 2 features were selected.\n")
  # If only one feature is selected, a boxplot is the most appropriate visualization
  plot_data_long <- data_full %>%
  dplyr::select(GroupName, all_of(selected_features)) %>%
  pivot_longer(
    cols = -GroupName,
    names_to = "Feature",
    values_to = "Intensity"
  ) %>%
  # A more robust way to reorder the features numerically for the plot
  mutate(Feature = fct_reorder(Feature, as.numeric(gsub("Cluster_", "", Feature))))

# --- Create the plot with improved aesthetics ---
boxplots_with_pvals <- ggplot(
  data = plot_data_long,
  aes(x = GroupName, y = Intensity, fill = GroupName)
) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.4, size = 1.5) +
  
  # Add Wilcoxon p-values to each panel
  stat_compare_means(
    method = "wilcox.test",
    label = "p.format", # e.g., "p = 0.04"
    label.x.npc = 0.5,  # Center the label
    size = 3.5
  ) +
  
  # Create a separate panel for each feature
  facet_wrap(~ Feature, scales = "free_y", ncol = 4) + # Arrange in 4 columns
  
  # --- EXPLICITLY DEFINE YOUR COLORS ---
  scale_fill_manual(
    name = "Group",
    values = c("Group 1 (pos)" = "darkgreen","Group 2 (neg)" = "darkorange") # Example: Blue and Yellow
    # â— IMPORTANT: Make sure these group names exactly match the levels in your `GroupName` column
  ) +
  
  labs(
    title = "Intensity Distribution of Boruta-Selected Features",
    subtitle = "P-values from Wilcoxon Rank-Sum Test",
    x = NULL, # Remove redundant "Group" label
    y = "Intensity (a.u)"
  ) +
  
  theme_bw(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5),
    strip.text = element_text(face = "bold"),
    legend.position = "none",
    axis.text.x = element_text(angle = 30, hjust = 1, vjust = 1)
  )

# This is the final line, which tells knitr to render the plot
boxplots_with_pvals
}
```
