---
title: "Comprehensive Analysis of GC-IMS Features for Cohort Classification"
author: "Deevanshi Walia"
date: "`r Sys.Date()`"
output:
  word_document:
    toc: yes
    toc_depth: 3
---

```{r setup, include=FALSE}
# This chunk sets up global options for the entire document.
# It will run, but neither the code nor its output will be shown.
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

# Load all libraries needed for the entire analysis
library(readr); library(dplyr); library(Boruta); library(MASS); library(caret);
library(ggplot2); library(GGally); library(vegan); library(pROC); library(e1071);
library(xgboost); library(tibble); library(tidyr); library(ggpubr); library(forcats);
library(knitr) 
```
# 1. Data Loading and Preparation

This section details the loading and initial preparation of the feature matrix that serves as the input for all subsequent analyses.

```{r load_data}
# Load libraries for this chunk
library(readr)
library(dplyr)
library(tibble)

# --- Configuration ---
feature_matrix_path <- "C:/Users/deevanshi.walia/Desktop/Try 2.0/Grouping acc to Jessy/Final_Analysis_Output_Peaks_Clustering_9_0.05/Feature_Matrices/feature_matrix_ROI_INTEGRATION.csv"

# --- Load and Prepare Data ---
data_full <- read_csv(feature_matrix_path, show_col_types = FALSE)
data_full$GroupName <- as.factor(data_full$GroupName)
if (any(is.na(data_full))) {
  data_full <- na.omit(data_full)
}
metadata <- data_full %>% dplyr::select(SampleName, GroupName)
feature_data <- data_full %>% dplyr::select(starts_with("Cluster_"))
```

The dataset was successfully loaded, containing `r nrow(feature_data)` samples and `r ncol(feature_data)` features.

# 2. Robust Feature Selection with Boruta

To identify a minimal, non-redundant, and predictively powerful subset of features, the Boruta algorithm was employed.

```{r run_boruta}
# Load libraries for this chunk
library(Boruta)
library(dplyr)
library(vegan)
library(MASS)
library(randomForest) 
library(e1071)
library(xgboost)
library(pROC)
library(caret)

# --- Run Boruta ---
set.seed(123)
boruta_result <- Boruta(
  x = feature_data,
  y = as.factor(metadata$GroupName),
  doTrace = 0,
  maxRuns = 700
)
selected_features <- getSelectedAttributes(boruta_result, withTentative = FALSE)
feature_data_best <- feature_data[, selected_features, drop = FALSE]
```

The Boruta algorithm confirmed **`r length(selected_features)`** features as statistically important. The selected features are: `r paste(selected_features, collapse=", ")`.

### Boruta Importance Plot


```{r plot_boruta_ggplot, fig.width=12, fig.height=7}
# Load the necessary libraries for this specific plot
library(ggplot2)
library(dplyr)
library(tidyr) # For the pivot_longer function
library(forcats) # For reordering factors

# --- Step 1: Extract the Importance History from the Boruta Object ---
# The raw data for the plot is stored inside the Boruta object.
# We convert it to a long-format "tidy" data frame for ggplot.
boruta_plot_data <- as.data.frame(boruta_result$ImpHistory) %>%
  # Remove the shadow feature columns from the main data for clarity
  dplyr::select(-matches("shadow")) %>%
  pivot_longer(
    cols = everything(),
    names_to = "Feature",
    values_to = "Importance"
  )

# --- Step 2: Get the Final Decision for Each Feature for Coloring ---
feature_decisions <- boruta_result$finalDecision %>%
  tibble::enframe(name = "Feature", value = "Decision")

# Join the decisions back to the plot data
boruta_plot_data <- boruta_plot_data %>%
  left_join(feature_decisions, by = "Feature") %>%
  # Filter out any rejected features to keep the plot clean
  filter(Decision != "Rejected") %>%
  # Reorder the features on the x-axis by their median importance
  mutate(Feature = fct_reorder(Feature, Importance, .fun = median))

# --- Step 3: Create the ggplot ---
boruta_ggplot <- ggplot(
  data = boruta_plot_data,
  aes(x = Feature, y = Importance, fill = Decision)
) +
  # Add the boxplots
  geom_boxplot(alpha = 0.8, outlier.shape = 21, outlier.size = 2) +
  
  # Define the colors for the decisions
  scale_fill_manual(
    name = "Boruta Decision",
    values = c("Confirmed" = "darkgreen", "Tentative" = "darkorange")
  ) +
  
  # Add a horizontal line at y=0 for reference
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  
  labs(
    title = "Boruta Feature Importance",
    subtitle = "Z-scores of feature importance across all iterations",
    x = "Features",
    y = "Importance (Z-score)"
  ) +
  
  coord_flip() +
  
  theme_bw(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5),
    legend.position = "bottom",
    axis.text.y = element_text(size = 10)
  )
boruta_ggplot
```

# 3. Statistical and Predictive Validation of the Selected Signature

The biomarker signature identified by Boruta was then subjected to a rigorous validation pipeline to assess its statistical significance and real-world predictive performance.

```{r model_gauntlet}
# --- This chunk now only performs calculations. Outputs are in later chunks. ---
library(vegan); library(MASS); library(randomForest); library(e1071);
library(xgboost); library(pROC); library(caret); library(tibble)

# --- PART C: STATISTICAL SIGNIFICANCE (PERMANOVA) ---
set.seed(123)
permanova_result <- adonis2(
  feature_data_best ~ GroupName,
  data = metadata,
  permutations = 999,
  method = "euclidean" # Use robust distance metric
)
permanova_table <- as.data.frame(permanova_result)

# --- PART D & E: MODEL GAUNTLET AND PERFORMANCE SUMMARY ---
num_samples <- nrow(feature_data_best); y_response <- as.factor(metadata$GroupName)
y_response_numeric <- as.numeric(y_response)-1; current_group_levels <- levels(y_response)
positive_class_name <- current_group_levels[2]
loocv_results <- data.frame(SampleName=metadata$SampleName, ActualGroup=y_response,
                            lda_scores=numeric(num_samples), rf_scores=numeric(num_samples),
                            svm_scores=numeric(num_samples), xgb_scores=numeric(num_samples))
for (i in 1:num_samples) {
  x_train <- as.matrix(feature_data_best[-i,,drop=F]); y_train <- y_response[-i]
  y_train_numeric <- y_response_numeric[-i]; x_test <- as.matrix(feature_data_best[i,,drop=F])
  
  lda_model<-lda(x=x_train, grouping=y_train); loocv_results$lda_scores[i]<-predict(lda_model,newdata=data.frame(x_test))$posterior[,positive_class_name]
  min_class_size<-min(table(y_train)); rf_model<-randomForest(x=x_train,y=y_train,ntree=500,sampsize=c(min_class_size,min_class_size))
  loocv_results$rf_scores[i]<-predict(rf_model,newdata=x_test,type="prob")[,positive_class_name]
  svm_model<-svm(x=x_train,y=y_train,probability=T,kernel="radial"); svm_pred<-predict(svm_model,newdata=data.frame(x_test),probability=T)
  loocv_results$svm_scores[i]<-attr(svm_pred,"probabilities")[,positive_class_name]
  dtrain<-xgb.DMatrix(data=x_train,label=y_train_numeric); dtest<-xgb.DMatrix(data=x_test)
  scale_pos_weight<-sum(y_train_numeric==0)/sum(y_train_numeric==1)
  params<-list(objective="binary:logistic",eval_metric="logloss",scale_pos_weight=scale_pos_weight)
  xgb_model<-xgboost(params=params,data=dtrain,nrounds=50,verbose=0)
  loocv_results$xgb_scores[i]<-predict(xgb_model,dtest)
}
roc_lda <- roc(response=loocv_results$ActualGroup, predictor=loocv_results$lda_scores, levels=current_group_levels)
roc_rf <- roc(response=loocv_results$ActualGroup, predictor=loocv_results$rf_scores, levels=current_group_levels)
roc_svm <- roc(response=loocv_results$ActualGroup, predictor=loocv_results$svm_scores, levels=current_group_levels)
roc_xgb <- roc(response=loocv_results$ActualGroup, predictor=loocv_results$xgb_scores, levels=current_group_levels)
model_performance_list <- list(); models_to_test <- list(LDA=roc_lda, `Random Forest`=roc_rf, SVM=roc_svm, XGBoost=roc_xgb); cm_list <- list() 
for (model_name in names(models_to_test)) {
  current_roc <- models_to_test[[model_name]]
  best_coords <- coords(current_roc, "best", ret="threshold", best.method="youden")
  predicted_classes <- ifelse(current_roc$predictor >= best_coords$threshold, current_group_levels[2], current_group_levels[1])
  cm_stats <- confusionMatrix(data=factor(predicted_classes, levels=current_group_levels), reference=current_roc$response, positive=positive_class_name)
  cm_list[[model_name]] <- cm_stats
  model_performance_list[[model_name]] <- c(AUC=auc(current_roc), cm_stats$overall["Accuracy"], cm_stats$byClass[c("Sensitivity","Specificity","Balanced Accuracy","Pos Pred Value","Neg Pred Value")])
}
performance_summary <- as.data.frame(do.call(rbind, model_performance_list)) %>%
  tibble::rownames_to_column("Model") %>%
  mutate(across(where(is.numeric), ~ round(., 4))) %>%
  dplyr::select(Model, AUC, Accuracy, `Balanced Accuracy`, Sensitivity, Specificity, `Pos Pred Value`, `Neg Pred Value`)
```

### 3.1. PERMANOVA Significance Test

```{r print_permanova_result}
# --- THIS IS THE KEY FIX FOR TABLES ---
# We wrap the output in kable() to create a beautiful table.
kable(permanova_table, caption = "PERMANOVA test results for the Boruta-selected feature signature.")
```

### 3.2. Comparative Model Performance

The table below summarizes the performance of four different classification models trained on the selected features and validated with Leave-One-Out Cross-Validation.

```{r print_performance_summary}
# Using kable() on the performance summary table
kable(performance_summary, caption = "Comparative performance metrics for all tested classifiers.")
```
### 3.2. ROC Curve and Best Model Analysis

```{r plot_roc_and_cm}
# --- Combined ROC Curve Plot (unchanged) ---
roc_list_for_plot <- list(LDA=roc_lda, RF=roc_rf, SVM=roc_svm, XGBoost=roc_xgb)
roc_plot_combined <- ggroc(roc_list_for_plot, size = 1) +
  annotate("segment", x = 1, xend = 0, y = 0, yend = 1, color="grey", linetype="dashed") +
  labs(title = "Comparative ROC Curves", color = "Model") +
  theme_minimal() + theme(plot.title = element_text(face = "bold", hjust = 0.5))

# This will embed the ROC plot in the final document
roc_plot_combined

# --- Best Model's Confusion Matrix (NEW, FORMATTED VERSION) ---
best_model_name <- performance_summary$Model[which.max(performance_summary$`Balanced Accuracy`)]
cat("\n--- Detailed Confusion Matrix for the Best Performing Model:", best_model_name, "---\n")

# Get the confusion matrix object for the best model
best_cm_object <- cm_list[[best_model_name]]

# --- 1. Create the main confusion matrix table ---
cm_table <- as.data.frame(best_cm_object$table)
colnames(cm_table) <- c("Prediction", "Reference", "Frequency")
cm_table_wide <- tidyr::pivot_wider(cm_table, names_from = "Reference", values_from = "Frequency")

# Use kable() to format it nicely
knitr::kable(
  cm_table_wide,
  caption = paste("Confusion Matrix for the", best_model_name, "Model")
)

# --- 2. Create a separate table for the key statistics ---
# Extract the key metrics we want to show
key_stats <- c(
  "Accuracy",
  "Kappa",
  "Balanced Accuracy",
  "Sensitivity",
  "Specificity",
  "Pos Pred Value",
  "Neg Pred Value"
)

# Combine overall and by-class stats
all_stats <- c(best_cm_object$overall, best_cm_object$byClass)

# Create a clean data frame of the stats
stats_to_display <- data.frame(
  Metric = key_stats,
  Value = round(all_stats[key_stats], 4) # Round to 4 decimal places
)

# Use kable() to format this table as well
knitr::kable(
  stats_to_display,
  caption = "Key Performance Statistics for the Best Model"
)
```


# 4. Characterization of the Biomarker Signature

This section visualizes the properties of the final features selected by Boruta.


### 4.1. Intensity Distributions

```{r plot_boxplots, fig.width=12, fig.height=8}
# --- Load all necessary libraries for this specific plot ---
library(ggplot2)
library(tidyr)
library(ggpubr)
library(forcats)
library(dplyr)

# --- Prepare data for plotting ---
# This ensures we are only working with the selected features
plot_data_long <- data_full %>%
  dplyr::select(GroupName, all_of(selected_features)) %>%
  pivot_longer(
    cols = -GroupName,
    names_to = "Feature",
    values_to = "Intensity"
  ) %>%
  # A more robust way to reorder the features numerically for the plot
  mutate(Feature = fct_reorder(Feature, as.numeric(gsub("Cluster_", "", Feature))))

# --- Create the plot with improved aesthetics ---
boxplots_with_pvals <- ggplot(
  data = plot_data_long,
  aes(x = GroupName, y = Intensity, fill = GroupName)
) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.4, size = 1.5) +
  
  # Add Wilcoxon p-values to each panel
  stat_compare_means(
    method = "wilcox.test",
    label = "p.format", # e.g., "p = 0.04"
    label.x.npc = 0.5,  # Center the label
    size = 3.5
  ) +
  
  # Create a separate panel for each feature
  facet_wrap(~ Feature, scales = "free_y", ncol = 4) + # Arrange in 4 columns
  
  # --- EXPLICITLY DEFINE YOUR COLORS ---
  scale_fill_manual(
    name = "Group",
    values = c("Group 1 (pos)" = "darkgreen","Group 2 (neg)" = "darkorange") # Example: Blue and Yellow
    # â— IMPORTANT: Make sure these group names exactly match the levels in your `GroupName` column
  ) +
  
  labs(
    title = "Intensity Distribution of Boruta-Selected Features",
    subtitle = "P-values from Wilcoxon Rank-Sum Test",
    x = NULL, # Remove redundant "Group" label
    y = "Intensity (a.u)"
  ) +
  
  theme_bw(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5),
    strip.text = element_text(face = "bold"),
    legend.position = "none",
    axis.text.x = element_text(angle = 30, hjust = 1, vjust = 1)
  )

# This is the final line, which tells knitr to render the plot
boxplots_with_pvals
```
